Lattice Boltzmann simulations of the dynamics of thin liquid films \\ \large{Gitter Boltzmann Simulationen der Dynamik von d\"unnen Fl\"ussigkeitsfilmen}

### 2024-01-16 chapters/Theory.tex

The starting point to study fluid dynamics is the Navier-Stokes equation~\cite{claude-louis-marie-henriMemoireLoisMouvement1827, stokesSteadyMotionIncompressible1848}, as motivated in Chap.~\ref{chapter:intro}.
Both Claude Navier in 1823 as well as Gorge G. Stokes in 1842 derived the momentum equation for a Newtonian liquid.
Historically these two were not the only ones. 
Siméon Denis Poisson and Barré de Saint-Venant came to the same conclusion around the same time~\cite{poissonNouvelleTheorieAction1831, de1843notea, jrHistoryAerodynamicsIts1998}.
But history is written and Navier as well as Stokes are identified with the derivation of the momentum of a fluid element.

\section{The Navier-Stokes equation}
\label{sec:navier_stokes_sec}
Mathematically the Navier-Stokes equation can be derived from one definition and two constraints.
The two constraints are mass and momentum conservation.
Mass conservation simply states that mass cannot be lost or created.
The same argument holds for the momentum.
By Noethers theorem every conserved quantity yields a symmetry and vice versa.
In this case conservation of momentum allows us to use the constraints on the solutions that translational symmetry requires.
Put simply a sliding drop experiment has to yield the same result in any laboratory on earths surface, independent of the exact location of the laboratory.
The last ingredient for the derivation is the definition of a fluid.
A fluid is ``\textit{a substance that flows and is not solid}`` according to the Cambridge Dictionary.
This definition can be refined, because a fluid is either a gas or a liquid that continuously deforms under shear stress, at least in the case of a Newtonian fluid.
And as such there is a linear relation between shear rate and shear velocity for a Newtonian fluid.
Clearly there are other liquids (which encloses granular systems to some extent) that exhibit different behaviour, for example toothpaste.
As long as there is no force applied to the tube the toothpaste does not flow.
Even the acceleration due gravity when pointing the tube towards the ground does not exert enough force to generate a flow.
In contrast to Newtonian liquids those liquids have a different response function to shear, they can be addressed as \textit{power law} fluids.
While the phenomenology of these fluids, e.g. shear thinning or thickening, is very interesting and important in many industrial applications it is beyond the scope of this work. 
That said, in Chap.~\ref{chapter:second_paper} an extension towards more complex flows is presented. 
Instead of however changing the shear response we model thermal fluctuations to approximate the stochastic thin film equation (STF).
In fact, the step to derive an extension to our approach are nicely summarized in Chap.~\ref{chapter:second_paper}. 
It should serve as a guideline for further features, such as viscoelasticity for non-Newtonian liquids~\cite{bouchutNewModelShallow2013}.

Knowing what a fluid is and taking into account the conservation of mass and momentum it is possible to derive the Navier-Stokes equation. 
By applying a continuity equation we assume that the fluid itself is a continuum and therefore ignoring the underlying molecular structure for now. 
This allows us to describe the fluids state with a density $\rho(\mathbf{x},t)$, instead of describing every atom or molecule individually.
Assuming an infinitesimal small volume of fluid $V_0$ its density can only change if there is a flow through the boundaries of the volume $\partial V_0$,
\begin{equation}\label{eq:cont_1}
    \partial_t\int_{V_0}\rho\diff V = -\oint_{\partial V_0} \rho\vec{u}\cdot\diff \vec{A}, 
\end{equation}
where $\vec{u}(\vec{x},t)$ is the fluid's velocity, A is the area that encloses the volume $V_0$ and $\diff\vec{A}$ are the normals pointing outwards of the volume.
Violations of this Eq.~(\ref{eq:cont_1}) are equivalent to generation or destruction of mass.

@@

For the momentum equation the same line of arguments is applied, however momentum can change for different reasons.
Momentum can for example be advected with the flow from or into the volume $V_0$.
Pressure on the other hand can be a source of momentum change and of course forces that are applied to the fluid are as well sources~\cite{krugerLatticeBoltzmannMethod2017}.
Collecting these terms and writing down an equation for the momentum we have,
\begin{equation}\label{eq:navier_stokes_1}
    \partial_t \int_{V_0}\rho\vec{u}\diff V = -\oint_{\partial V_0}(\rho\vec{u}\vec{u})\cdot\diff\vec{A} - \oint_{\partial V_0} p\diff\vec{A} + \int_{V_0} \vec{F}\diff V, 
\end{equation}
where $\vec{u}\vec{u}$ is used to denote the outer product
\begin{equation}
    \vec{u}\vec{u} = \begin{bmatrix}
    u_x u_x & u_x u_y & u_x u_z \\
    u_y u_x & u_y u_y & u_y u_z \\
    u_z u_x & u_z u_y & u_z u_z
    \end{bmatrix} .
\end{equation}
Using the divergence theorem again to reformulate everything in terms of volume integrals and dropping the integration the above equation becomes
\begin{equation}\label{eq:navier_stokes_2}
    \partial_t(\rho\vec{u}) = -\vec{\nabla}\cdot(\rho\vec{u}\vec{u}) - \vec{\nabla} p + \vec{F}.
\end{equation}
This however is not yet the Navier-Stokes equation, in fact this is the Euler equation~\cite{batchelorIntroductionFluidDynamics1967}.
In the derivation there is one term missing, because the momentum of a fluid can also change due to internal friction, or put simply, due to viscosity.
These viscous contributions are collected in the viscous stress tensor $\hat{\sigma}$ as
\begin{equation}\label{eq:stress_tens}
    \hat{\sigma}_{\alpha\beta} = \mu\left(\frac{\partial u_{\alpha}}{\partial x_{\beta}} + \frac{\partial u_{\beta}}{\partial x_{\alpha}}\right) + \lambda\delta_{\alpha\beta}\frac{\partial u_{\gamma}}{\partial x_{\gamma}},
\end{equation}
This tensor introduces a lot of the complexity of the Navier-Stokes equation as in principle neither $\mu$ nor $\lambda$ need to be constants and in fact they are tensors themselves.

@@

Inserting this constraint Eq.~(\ref{eq:incomp}) into Eq.~(\ref{eq:navier_stokes_3}) yields therefore the Navier-Stokes equation for an incompressible fluid as
\begin{equation}\label{eq:navier_stokes_4}
    \rho(\partial_t\vec{u} + \vec{\nabla}\cdot\vec{u}\vec{u}) = - \vec{\nabla} p + \mu\Delta\vec{u} + \vec{F},
\end{equation}
where the symbol $\Delta$ denotes the laplace operator ($\Delta f = \vec{\nabla}\cdot\vec{\nabla} f$).
The left hand side can be rewritten using the material derivative
\begin{equation}\label{eq:mat_dev}
    D_t = \partial_t + \vec{u}\cdot\vec{\nabla}, 
\end{equation}
which is a convenient measure for following a fluid element instead of a single point.
Inserting this, Eq.~(\ref{eq:navier_stokes_4}) can be written as
\begin{equation}\label{eq:navier_stokes_fin}
    \rho D_t \vec{u} =  - \vec{\nabla} p + \mu\Delta\vec{u} + \vec{F}.
\end{equation}
Eqs.~(\ref{eq:cont_3}, \ref{eq:navier_stokes_4}) cannot be solved because there are too few equations for too many unknowns.
So far there are five unknowns ($\rho, \mathbf{u}, p$) for four equations.
The clever way out of this misery is to use an equation of state that relates pressure, density, temperature, energy and entropy.
Especially the relation between pressure, density and temperature helps in closing the system of equations.
To define an equation of state is far from trivial.
It requires a thermodynamically consistent approach that is based on observations or theoretical assumptions.
The equation of state for an ideal gas was born from theoretical assumptions, e.g. non-interacting gas molecules, and reads~\cite{schwablStatisticalMechanics2006}
\begin{equation}\label{eq:eq_of_state}
    p = \rho R T,
\end{equation}
where $R$ is the specific gas constant and $T$ is the temperature.
Having a well defined state in the thermodynamical sense allows to relate pressure and density and therefore to close the system of equations.
The equation of state for an ideal gas, see Eq.~(\ref{eq:eq_of_state}), will be revisited in Chap.~\ref{chapter:method} where we discuss the lattice Boltzmann method.

@@

\section{Shallow Water Theory}
\label{sec:theory_shallow_water}
Solving the Navier-Stokes equation for the problems discussed in Chaps.~\ref{chapter:fourth_paper}-\ref{chapter:third_paper} is not needed.
Looking at Figs.~\ref{fig:examples_intro}-\ref{fig:morph_transition} one can identify that the height or thickness of the fluid layer is much smaller than the horizontal extension of the substrate.
This statement together with some more approximations allows to derive another system from the Navier-Stokes equation and the continuity equation which is called the shallow water or Saint-Venant equations~\cite{bTheorieMouvementNonpermanent1871}.

@@

The task now is to make, Eq.~(\ref{eq:navier_stokes_4}), aware about these constraints.
Although this sounds straightforward partial differential equations are complex. 
They are one of the topics in mathematics where researchers are satisfied with the irritating result that an equation cannot be solved, an undesired outcome for most Physicists.
Nevertheless, to come up with consistent strategies to solve them one can classify them, e.g. hyperbolic, parabolic, linear, nonlinear, initial value problem or boundary value problem.
There is in fact overlap between those terms, however to keep things simple it is sufficient to talk about initial or boundary value problems.
The former can be solved knowing the initial configuration which then defines the evolution, e.g. the initial coordinates and velocity of a pendulum. 
For the latter ones it is mandatory to know the solution at the boundaries. 
If we know the current temperature distribution on a metallic plate and the heat flux on the boundaries then we are able to calculate the temperature field on that plate at later times.
The shallow water system can be derived from the Navier-Stokes equation using both well defined boundary conditions and system specific approximations.
For example the statement \textit{pressure free surface} translates to the fact that there is a boundary condition requiring $p = 0$ at $z = \eta$. 
Therefore no pressure is acting on liquids interface.

@@

In Eqs.~(\ref{eq:vel_scaling}-\ref{eq:scaling_thin_layer}) there is no distinct difference between both horizontal dimensions $(x,y)$. 
For the sake of clarity and to simplify the derivation of the shallow water model we neglect the $y$-components.
Of course, to study large oceanic systems or Coriolis forces one needs both horizontal dimensions, see e.g.~\cite{dellarShallowWaterEquations2005, marcheDerivationNewTwodimensional2007}, but for the following derivation it is sufficient to use only one.
Furthermore we use Eq.~(\ref{eq:long_wave_scaling}), which represents long wave theory or scaling. 
We make use of this approach once more in the derivation of the thin film equation.
The benefit of this choice is the smoothing effect on variations which makes them more gradual. 
But not only the gradients in horizontal dimensions are multiplied with $\varepsilon^{-1}$.
The time is scaled with $\varepsilon^{-1}$ as well which smears the dynamics.
Inserting Eqs.~(\ref{eq:vel_scaling},\ref{eq:long_wave_scaling}) into Eq.~(\ref{eq:cont_3}) yields
\begin{equation}\label{eq:rescaled_cont}
    \tilde{\nabla}\cdot\tilde{\mathbf{u}} = \partial_{\tilde{x}} \tilde{u}_x + \partial_{\tilde{z}} \tilde{u}_z  = 0,
\end{equation}
where two dimensional vectors, such as $\mathbf{u} = (u_x, u_z)$ are bold while three dimensional ones are denoted with an arrow symbol.

Scaling analysis and renormalization are key concepts in modern physics~\cite{yakhotRenormalizationGroupAnalysis1986, IntroductionQuantumField2018}.
One of the key concepts of this idea are symmetries, and therefore transformations under which a given state does not change.
Fluid dynamics to some extent admits a conformal symmetry.
The governing equations describe the flow through a nanotube as well as the circular currents in oceans~\cite{secchiMassiveRadiusdependentFlow2016, marcheDerivationNewTwodimensional2007}.
Some features however such as the drag on an airfoil do not simply scale with the size of the system.
Therefore while a miniaturised model may create enough lift force, simply scaling up the size is not sufficient to ensure enough lift.  
In the following it will be helpful to introduce dimensionless numbers~\cite{ruzickaDimensionlessNumbers2008,neversFluidMechanicsChemical2019}.
These numbers have a long history in fluid dynamics.
The Reynolds number for example is a measure if a flow is turbulent or not, it depends only on proportions and not on an exact experiment.
With the knowledge of these number comparing experiments is straightforward, even if the experiments differ in many aspects.
The two dimensionless numbers of interest for the shallow water theory are the Reynolds number and the Froud number
\begin{equation}\label{eq:Re_and_Fr}
    Re = \frac{\rho L v_0}{\mu},\quad Fr = \frac{v_0}{\sqrt{g L}},
\end{equation}
where $L$ is a characteristic length scale of the problem, e.g. the mean water depth, $v_0$ and $g$ are a characteristic velocity and the gravitational acceleration ($g = 9.81 m/s^2$), respectively.
A characteristic velocity can be for example the inflow velocity from a boundary, or the speed of a certain wave type such as a tsunami.
Both numbers measure the magnitude of certain effects, in the case of the Reynolds number the inertial forces are compared against viscosity.
The Froud number is a measure to compare kinetic ($v_0$) with potential energy ($g L$).

In the next step both $Fr$ and $Re$ are used to make Eq.~(\ref{eq:navier_stokes_fin}) dimensionless.
For convenience the vector equation is split into two respective equations with gravity being the only force present.
Writing the full equation with all components yields
\begin{align}\label{eq:2d_RNS}
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_z u_x &= -\partial_x p + \frac{1}{Re}\Delta u_x ,\\
    \partial_t u_z + u_z\partial_x u_x + u_z\partial_z u_z &= -\partial_z p - \frac{1}{Fr^2} + \frac{1}{Re}\Delta u_z ,
\end{align}
with the two dimensional Laplacian $\Delta = (\partial_x^2 + \partial_z^2)$.
Using long wave scaling, Eq.~(\ref{eq:long_wave_scaling}) and the rescaled length and time variables (e.g. $\tilde{x}$),
\begin{align}\label{eq:2d_RNS_eps}
    \varepsilon(\partial_{\tilde{t}} \tilde{u}_x + \tilde{u}_x \partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z \partial_{\tilde{z}} \tilde{u}_x ) &= -\varepsilon\partial_{\tilde{x}} \tilde{p} + \frac{1}{Re}(\varepsilon^2 \partial_{\tilde{x}}^2 \tilde{u}_x + \partial_{\tilde{z}}^2 \tilde{u}_x ) ,\\
    \varepsilon^2(\partial_{\tilde{t}} \tilde{u}_z + \tilde{u}_z\partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z\partial_{\tilde{z}} \tilde{u}_z) &= -\partial_{\tilde{z}}\tilde{p} - \frac{1}{Fr^2} + \frac{\varepsilon}{Re}(\varepsilon^2 \partial_{\tilde{x}}^2 \tilde{u}_z + \partial_{\tilde{z}}^2 \tilde{u}_z ) .
\end{align}
We now assume that terms with higher orders in $\varepsilon$ are negligible and therefore $O(\varepsilon^2) \rightarrow 0$, collecting only terms with up to $O(\varepsilon)$
\begin{align}\label{eq:2d_RNS_linear}
    \partial_{\tilde{t}} \tilde{u}_x + \tilde{u}_x\partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z\partial_{\tilde{z}} \tilde{u}_x &= -\partial_{\tilde{x}} \tilde{p} + \frac{\partial_{\tilde{z}}^2\tilde{u}_x}{\varepsilon Re}, \\
    0 &= -\partial_{\tilde{z}}\tilde{p} - \frac{1}{Fr^2} + \frac{\varepsilon\partial_{\tilde{z}}^2 \tilde{u}_z}{Re} .
\end{align}
Interestingly in the first of the two above equations the term associated to the horizontal viscosity is $O(\varepsilon^2)$ and can be dropped.
On the other hand with the assumption of the long wave scaling the second of the above equations yields the hydrostatic pressure, taking only $O(1)$ terms
\begin{equation}\label{eq:hydro_static}
    \partial_{\tilde{z}}\tilde{p} = -\frac{1}{Fr^2},
\end{equation}
which can be readily integrated, given $Fr$ is a constant, using the upper bound at the liquid air interface $\eta(x)$ 
\begin{equation}\label{eq:hydro_static_int}
    \tilde{p} = \frac{1}{Fr^2}(\tilde{\eta} - \tilde{z}).   
\end{equation}
This implies that $\partial_{\tilde{x}}p$ does not depend on $\tilde{z}$ and therefore states that the pressure gradient is conserved along z.

@@

The first one is Boussinesq coefficient $\beta$ which is defined by the equation

@@

and the second one is the friction or the shear stress at the bottom of the fluid column $\tau_b$

@@

Both $\beta$ as well as $\tau_b$ do depend on $u$, therefore to know their exact contribution one needs to know about the flow.

@@

A word of caution when using $\tau_b$.
Friction forces in real experiments are very dependent on the actual boundary condition at $z = b$.
Both the topography and the specific type of material at the bottom will influence the measured value of $\tau_b$. 
By modelling a dissipative force such as friction we make an error, sometimes this is error is negligible.
It is therefore often simpler and more accurate to use data from empirical studies or real world experiments.
Within the shallow water literature an often encountered coefficient is the one of Manning, the empirical law that connects flow velocity with friction can be found in Ref.~\cite{f.asceOpenChannelHydraulics2021}.

@@

In the derivation of the shallow water equations we encountered an interface, but it's shape was solely determined by the hydrostatic pressure.

@@

The disjoining pressure ($\Pi(h)$) on the other hand takes into account the inter-molecular interactions that can arise, e.g. the interaction between solid substrate and film~\cite{crasterDynamicsStabilityThin2009, moultonEffectDisjoiningPressure2013}.

@@

where different $\gamma$'s denote the different surface energies, e.g. $\gamma_{vs}$ is the surface energy between the vapour and the solid phases.

@@

While we can easily formulate Eq.~(\ref{eq:surface_energies_contact_angle}) it is almost impossible to measure the different $\gamma$s experimentally.

>>>>

\section{The Navier-Stokes equation}
\label{sec:navier_stokes_sec}
The starting point to study fluid dynamics is the Navier-Stokes equation~\cite{claude-louis-marie-henriMemoireLoisMouvement1827, stokesSteadyMotionIncompressible1848}, as motivated in Chap.~\ref{chapter:intro}.
Both Claude Navier in 1823 as well as Gorge G. Stokes in 1842 derived the momentum equation for a Newtonian liquid.
Historically these two were not the only ones. 
Siméon Denis Poisson and Barré de Saint-Venant came to the same conclusion around the same time~\cite{poissonNouvelleTheorieAction1831, de1843notea, jrHistoryAerodynamicsIts1998}.
But history is written and Navier as well as Stokes are identified with the derivation of the momentum of a fluid element.

Mathematically the Navier-Stokes equation can be derived from one definition and two constraints.
The two constraints are mass and momentum conservation.
Mass conservation simply states that mass cannot be lost or created.
The same argument holds for the momentum.
By Noethers theorem every conserved quantity yields a symmetry and vice versa.
In this case conservation of momentum allows us to use the constraints on the solutions that translational symmetry requires.
Put simply a sliding drop experiment has to yield the same result in any laboratory on earths surface, independent of the exact location of the laboratory.
The last ingredient for the derivation is the definition of a fluid.
A fluid is ``\textit{a substance that flows and is not solid}`` according to the Cambridge Dictionary.
This definition can be refined, because a fluid is either a gas or a liquid that continuously deforms under shear stress, at least in the case of a Newtonian fluid.
And as such there is a linear relation between shear rate and shear velocity for a Newtonian fluid.
Clearly there are other liquids (which encloses granular systems to some extent) that exhibit different behaviour, for example toothpaste.
As long as there is no force applied to the tube the toothpaste does not flow.
Even the acceleration due to gravity when pointing the tube towards the ground does not exert enough force to generate a flow.
In contrast to Newtonian liquids those liquids have a different response function to shear, they can be addressed as \textit{power law} fluids.
While the phenomenology of these fluids, e.g., shear thinning or thickening, is very interesting and important in many industrial applications it is beyond the scope of this work. 
That said, in Chap.~\ref{chapter:second_paper} an extension towards more complex flows is presented. 
Instead of however changing the shear response we model thermal fluctuations to approximate the stochastic thin film equation (STF).
In fact, the step to derive an extension to our approach are nicely summarized in Chap.~\ref{chapter:second_paper}. 
It should serve as a guideline for further features, such as viscoelasticity for non-Newtonian liquids~\cite{bouchutNewModelShallow2013}.

Knowing what a fluid is and taking into account the conservation of mass and momentum it is possible to derive the Navier-Stokes equation. 
By applying a continuity equation we assume that the fluid itself is a continuum and therefore ignore the underlying molecular structure for now. 
This allows us to describe the fluid state with a density $\rho(\mathbf{x},t)$, instead of describing every atom or molecule individually.
Assuming an infinitesimal small volume of fluid $V_0$ its density can only change if there is a flow through the boundaries of the volume $\partial V_0$,
\begin{equation}\label{eq:cont_1}
    \partial_t\int_{V_0}\rho\diff V = -\oint_{\partial V_0} \rho\vec{u}\cdot\diff \vec{A}, 
\end{equation}
where $\vec{u}(\vec{x},t)$ is the fluid's velocity, A is the area that encloses the volume $V_0$ and $\diff\vec{A}$ are the normals pointing outwards of the volume.
Violations of this Eq.~(\ref{eq:cont_1}) are equivalent to generation or destruction of mass.

@@

For the momentum equation the same line of arguments is applied, however momentum can change for different reasons.
Momentum can for example be advected with the flow from or into the volume $V_0$.
Pressure on the other hand can be a source of momentum change and of course forces that are applied to the fluid are as well sources~\cite{krugerLatticeBoltzmannMethod2017}.
Collecting these terms and writing down an equation for the momentum we have,
\begin{equation}\label{eq:navier_stokes_1}
    \partial_t \int_{V_0}\rho\vec{u}\diff V = -\oint_{\partial V_0}(\rho\vec{u}\vec{u})\cdot\diff\vec{A} - \oint_{\partial V_0} p\diff\vec{A} + \int_{V_0} \vec{F}\diff V, 
\end{equation}
where $\vec{u}\vec{u}$ is used to denote the outer product
\begin{equation}
    \vec{u}\vec{u} = \begin{bmatrix}
    u_x u_x & u_x u_y & u_x u_z \\
    u_y u_x & u_y u_y & u_y u_z \\
    u_z u_x & u_z u_y & u_z u_z
    \end{bmatrix} .
\end{equation}
Using the divergence theorem again to reformulate everything in terms of volume integrals and dropping the integration the above equation becomes
\begin{equation}\label{eq:navier_stokes_2}
    \partial_t(\rho\vec{u}) = -\vec{\nabla}\cdot(\rho\vec{u}\vec{u}) - \vec{\nabla} p + \vec{F}.
\end{equation}
This however is not yet the Navier-Stokes equation, in fact this is the Euler equation~\cite{batchelorIntroductionFluidDynamics1967}.
In the derivation there is one term missing, because the momentum of a fluid can also change due to internal friction, or put simply, due to viscosity.
These viscous contributions are collected in the viscous stress tensor $\hat{\sigma}$ as
\begin{equation}\label{eq:stress_tens}
    \hat{\sigma}_{\alpha\beta} = \mu\left(\frac{\partial u_{\alpha}}{\partial x_{\beta}} + \frac{\partial u_{\beta}}{\partial x_{\alpha}}\right) + \lambda\delta_{\alpha\beta}\frac{\partial u_{\gamma}}{\partial x_{\gamma}}.
\end{equation}
This tensor introduces a lot of the complexity to the Navier-Stokes equation as in principle neither $\mu$ nor $\lambda$ need to be constants and in fact they are tensors themselves.

@@

Inserting this constraint Eq.~(\ref{eq:incomp}) into Eq.~(\ref{eq:navier_stokes_3}) yields therefore the Navier-Stokes equation for an incompressible fluid as
\begin{equation}\label{eq:navier_stokes_4}
    \rho(\partial_t\vec{u} + \vec{\nabla}\cdot\vec{u}\vec{u}) = - \vec{\nabla} p + \mu\Delta\vec{u} + \vec{F},
\end{equation}
where the symbol $\Delta$ denotes the Laplace operator ($\Delta f = \vec{\nabla}\cdot\vec{\nabla} f$).
The left hand side can be rewritten using the material derivative
\begin{equation}\label{eq:mat_dev}
    D_t = \partial_t + \vec{u}\cdot\vec{\nabla}, 
\end{equation}
which is a convenient measure for following a fluid element instead of a single point.
Inserting this, Eq.~(\ref{eq:navier_stokes_4}) can be written as
\begin{equation}\label{eq:navier_stokes_fin}
    \rho D_t \vec{u} =  - \vec{\nabla} p + \mu\Delta\vec{u} + \vec{F}.
\end{equation}
Eqs.~(\ref{eq:cont_3}, \ref{eq:navier_stokes_4}) cannot be solved because there are too few equations for too many unknowns.
So far there are five unknowns ($\rho, \mathbf{u}, p$) for four equations.
The clever way out of this misery is to use an equation of state that relates pressure, density, temperature, energy and entropy.
Especially the relation between pressure, density and temperature helps in closing the system of equations.
To define an equation of state is far from trivial.
It requires a thermodynamically consistent approach that is based on observations or theoretical assumptions.
The equation of state for an ideal gas was born from theoretical assumptions, e.g., non-interacting gas molecules, and reads~\cite{schwablStatisticalMechanics2006}
\begin{equation}\label{eq:eq_of_state}
    p = \rho R T,
\end{equation}
where $R$ is the specific gas constant and $T$ is the temperature.
Having a well defined state in the thermodynamical sense allows to relate pressure and density and therefore to close the system of equations.
The equation of state for an ideal gas, see Eq.~(\ref{eq:eq_of_state}), will be revisited in Chap.~\ref{chapter:method} where we discuss the lattice Boltzmann method.

@@

\section{Shallow water theory}
\label{sec:theory_shallow_water}
Solving the Navier-Stokes equation for the problems discussed in Chaps.~\ref{chapter:fourth_paper}-\ref{chapter:third_paper} is not needed.
Looking at Figs.~\ref{fig:examples_intro}-\ref{fig:morph_transition} one can identify that the height or thickness of the fluid layer is much smaller than the horizontal extension of the substrate.
This statement together with some more approximations allows to derive another system from the Navier-Stokes equation and the continuity equation which is called the shallow water or Saint-Venant equations~\cite{bTheorieMouvementNonpermanent1871}.

@@

The task now is to make, Eq.~(\ref{eq:navier_stokes_4}), aware about these constraints.
Although this sounds straightforward partial differential equations are complex. 
They are one of the topics in mathematics where researchers are satisfied with the irritating result that an equation cannot be solved, an undesired outcome for most Physicists.
Nevertheless, to come up with consistent strategies to solve them one can classify them, e.g., hyperbolic, parabolic, linear, nonlinear, initial value problem or boundary value problem.
There is in fact overlap between those terms, however to keep things simple it is sufficient to talk about initial or boundary value problems.
The former can be solved knowing the initial configuration which then defines the evolution, e.g., the initial coordinates and velocity of a pendulum. 
For the latter ones it is mandatory to know the solution at the boundaries. 
If we know the current temperature distribution on a metallic plate and the heat flux on the boundaries then we are able to calculate the temperature field on that plate at later times.
The shallow water system can be derived from the Navier-Stokes equation using both well defined boundary conditions and system specific approximations.
For example the statement \textit{pressure free surface} translates to the fact that there is a boundary condition requiring $p = 0$ at $z = \eta$. 
Therefore no pressure is acting on liquids interface.

@@

In Eqs.~(\ref{eq:vel_scaling}-\ref{eq:scaling_thin_layer}) there is no distinct difference between both horizontal dimensions $(x,y)$. 
For the sake of clarity and to simplify the derivation of the shallow water model we neglect the $y$-components.
Of course, to study large oceanic systems or Coriolis forces one needs both horizontal dimensions, see e.g.,~\cite{dellarShallowWaterEquations2005, marcheDerivationNewTwodimensional2007}, but for the following derivation it is sufficient to use only one.
Furthermore we use Eq.~(\ref{eq:long_wave_scaling}), which represents long wave theory or scaling. 
We make use of this approach once more in the derivation of the thin film equation.
The benefit of this choice is the smoothing effect on variations which makes them more gradual. 
But not only the gradients in horizontal dimensions are multiplied with $\varepsilon^{-1}$.
The time is scaled with $\varepsilon^{-1}$ as well which smears the dynamics.
Inserting Eqs.~(\ref{eq:vel_scaling},\ref{eq:long_wave_scaling}) into Eq.~(\ref{eq:cont_3}) yields
\begin{equation}\label{eq:rescaled_cont}
    \tilde{\nabla}\cdot\tilde{\mathbf{u}} = \partial_{\tilde{x}} \tilde{u}_x + \partial_{\tilde{z}} \tilde{u}_z  = 0,
\end{equation}
where two dimensional vectors, such as $\mathbf{u} = (u_x, u_z)$ are bold while three dimensional ones are denoted with an arrow symbol.

Scaling analysis and renormalization are key concepts in modern physics~\cite{yakhotRenormalizationGroupAnalysis1986, IntroductionQuantumField2018}.
One of the key concepts of this idea are symmetries, and therefore transformations under which a given state does not change.
Fluid dynamics to some extent admits a conformal symmetry.
The governing equations describe the flow through a nanotube as well as the circular currents in oceans~\cite{secchiMassiveRadiusdependentFlow2016, marcheDerivationNewTwodimensional2007}.
Some features however such as the drag on an airfoil do not simply scale with the size of the system.
Therefore while a miniaturised model may create enough lift force, simply scaling up the size is not sufficient to ensure enough lift.  
In the following it will be helpful to introduce dimensionless numbers~\cite{ruzickaDimensionlessNumbers2008,neversFluidMechanicsChemical2019}.
These numbers have a long history in fluid dynamics.
The Reynolds number for example is a measure if a flow is turbulent or not, it depends only on proportions and not on an exact experiment.
With the knowledge of these number comparing experiments is straightforward, even if the experiments differ in many aspects.
The two dimensionless numbers of interest for the shallow water theory are the Reynolds number and the Froude number
\begin{equation}\label{eq:Re_and_Fr}
    Re = \frac{\rho L v_0}{\mu},\quad Fr = \frac{v_0}{\sqrt{g L}},
\end{equation}
where $L$ is a characteristic length scale of the problem, e.g., the mean water depth, $v_0$ and $g$ are a characteristic velocity and the gravitational acceleration ($g = 9.81 m/s^2$), respectively.
A characteristic velocity can be for example the inflow velocity from a boundary, or the speed of a certain wave type such as a tsunami.
Both numbers measure the magnitude of certain effects, in the case of the Reynolds number the inertial forces are compared against viscosity.
The Froude number is a measure to compare kinetic ($v_0$) with potential energy ($g L$).

In the next step both $Fr$ and $Re$ are used to make Eq.~(\ref{eq:navier_stokes_fin}) dimensionless.
For convenience the vector equation is split into two respective equations with gravity being the only force present.
Writing the full equation with all components yields
\begin{align}\label{eq:2d_RNS}
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_z u_x &= -\partial_x p + \frac{1}{Re}\Delta u_x ,\\
    \partial_t u_z + u_z\partial_x u_x + u_z\partial_z u_z &= -\partial_z p - \frac{1}{Fr^2} + \frac{1}{Re}\Delta u_z ,
\end{align}
with the two dimensional Laplacian $\Delta = (\partial_x^2 + \partial_z^2)$.
Using long wave scaling, Eq.~(\ref{eq:long_wave_scaling}) and the rescaled length and time variables (e.g., $\tilde{x}$),
\begin{align}\label{eq:2d_RNS_eps}
    \varepsilon(\partial_{\tilde{t}} \tilde{u}_x + \tilde{u}_x \partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z \partial_{\tilde{z}} \tilde{u}_x ) &= -\varepsilon\partial_{\tilde{x}} \tilde{p} + \frac{1}{Re}(\varepsilon^2 \partial_{\tilde{x}}^2 \tilde{u}_x + \partial_{\tilde{z}}^2 \tilde{u}_x ) ,\\
    \varepsilon^2(\partial_{\tilde{t}} \tilde{u}_z + \tilde{u}_z\partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z\partial_{\tilde{z}} \tilde{u}_z) &= -\partial_{\tilde{z}}\tilde{p} - \frac{1}{Fr^2} + \frac{\varepsilon}{Re}(\varepsilon^2 \partial_{\tilde{x}}^2 \tilde{u}_z + \partial_{\tilde{z}}^2 \tilde{u}_z ) .
\end{align}
We now assume that terms with higher orders in $\varepsilon$ are negligible and therefore $O(\varepsilon^2) \rightarrow 0$, collecting only terms with up to $O(\varepsilon)$
\begin{align}\label{eq:2d_RNS_linear}
    \partial_{\tilde{t}} \tilde{u}_x + \tilde{u}_x\partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z\partial_{\tilde{z}} \tilde{u}_x &= -\partial_{\tilde{x}} \tilde{p} + \frac{\partial_{\tilde{z}}^2\tilde{u}_x}{\varepsilon Re}, \\
    0 &= -\partial_{\tilde{z}}\tilde{p} - \frac{1}{Fr^2} + \frac{\varepsilon\partial_{\tilde{z}}^2 \tilde{u}_z}{Re} .
\end{align}
Interestingly in the first of the two above equations the term associated to the horizontal viscosity is $O(\varepsilon^2)$ and can be dropped.
On the other hand with the assumption of the long wave scaling the second of the above equations yields the hydrostatic pressure, taking only $O(1)$ terms
\begin{equation}\label{eq:hydro_static}
    \partial_{\tilde{z}}\tilde{p} = -\frac{1}{Fr^2},
\end{equation}
which can be readily integrated, given $Fr$ is a constant, using the upper bound at the liquid air interface $\eta(x)$ 
\begin{equation}\label{eq:hydro_static_int}
    \tilde{p} = \frac{1}{Fr^2}(\tilde{\eta} - \tilde{z}).   
\end{equation}
This implies that $\partial_{\tilde{x}}p$ does not depend on $\tilde{z}$ and therefore states that the pressure gradient is conserved along z.

@@

The first one is the Boussinesq coefficient $\beta$ which is defined by the equation

@@

and the second one is the friction or the shear stress at the bottom of the fluid column

@@

Both $\beta$ as well as $\tau_b$ do depend on $u$. 
Therefore to know their exact contribution one needs to know about the flow.

@@

A word of caution is needed when using $\tau_b$.
Friction forces in real experiments are dependent on the actual boundary condition at $z = b$.
Both the topography and the specific type of material at the bottom will influence the measured value of $\tau_b$. 
By modelling a dissipative force such as friction we make an error, and a priori we do not know if that error is large or small.
It is therefore often simpler and more accurate to use data from empirical studies or real world experiments.
Within the shallow water literature an often encountered coefficient is the one of Manning, the empirical law that connects flow velocity with friction can be found in Ref.~\cite{f.asceOpenChannelHydraulics2021}.

@@

In the derivation of the shallow water equations we encountered an interface, but its shape was solely determined by the hydrostatic pressure.

@@

The disjoining pressure ($\Pi(h)$) on the other hand takes into account the inter-molecular interactions that can arise, e.g., the interaction between solid substrate and film~\cite{crasterDynamicsStabilityThin2009, moultonEffectDisjoiningPressure2013}.

@@

where different $\gamma$'s denote the different surface energies, e.g., $\gamma_{vs}$ is the surface energy between the vapour and the solid phases.

@@

While we can easily formulate Eq.~(\ref{eq:surface_energies_contact_angle}) it is almost impossible to measure the different $\gamma$'s experimentally.
What can be measured in experiments on the other hand are the spreading parameter $S$ and/or the Hamaker constant~\cite{hamakerLondonVanWaals1937, bonnWettingSpreading2009, diezMetallicthinfilmInstabilitySpatially2016, peschkaSignaturesSlipDewetting2019} from atomic force microscopy (AFM) or dielectric response spectra~\cite{bergstromHamakerConstantsInorganic1997}. 
